---
output:
  github_document:
    toc: true
    toc_depth: 3
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# `ddeco`: Detailed decompositions of between-group differences in R

## Overview

**ddeco** implements the [Oaxaca (1973)](https://www.jstor.org/stable/2525981)-[Blinder (1973)](https://www.jstor.org/stable/144855) decomposition method and its generalizations that decompose differences in distributional statistics beyond the mean.

`ob_deco()` decomposes differences in the mean outcome between two groups into one 
part explained by different covariates (composition effect) and into 
another part due to differences in the way covariates are linked to the 
outcome variable (wage structure effect). The function further divides the two
effects into the contribution of each covariate and allows for weighted 'doubly
robust' decompositions. For distributional statistics beyond the mean,
the function performs the RIF decomposition proposed by [Firpo, Fortin, and Lemieux (2018)]( https://doi.org/10.3390/econometrics6020028). 

`dfl_deco()` divides differences in distributional statistics into an
composition effect and a wage structure effect using inverse probability
weighting as introduced by  [DiNardo, Fortin, and Lemieux (1996)](https://www.jstor.org/stable/2171954). The function also allows
to sequentially decompose the composition effect into the contribution of
single covariates.

The package contains generic summary, print and plot functions for the results
and computes bootstrapped standard errors. This documentation provides a brief 
overview of the functions implemented in the package. For a more detailed 
discussion of decomposition methods including their respective assumptions and 
limitations, refer to [Fortin, Lemieux, and Firpo (2011)](https://economics.ubc.ca/wp-content/uploads/sites/38/2013/05/pdf_paper_nicole-fortin-decomposition-methods.pdf) and Firpo et al. (2018). 

## Installation 

You can either install the CRAN version of `ddeco`
```{r, eval=FALSE}
install.packages("ddeco")
```

or the latest development version from GitHub:
```{r , eval=FALSE}
devtools::install_github("samumei/ddeco")
```


# New

## Background

### Oacaxa-Blinder decomposition

The original decomposition method introduced by Oaxaca (1973) and Blinder (1973) 
decomposes the difference in the mean of an outcome variable (e.g., hourly wages)
between two groups $g = 0, 1$ into a part explained by differences in the 
mean of the covariates (e.g., educational level or experience) and another part
due to different linear regression coefficients (e.g., returns to education) 
linking covariates to the outcome. 

The method linearly models the relationship between the outcome $Y$ and covariates
$X$
$$Y_{g,i} = \beta_{g,0} + \sum^K_{k=1}X_{k,i}\beta_{g,k} + \varepsilon_{g,i},$$
where $\beta_{g,0}$ is the intercept and $\beta_{g,k}$ are the slope coefficients of
covariates $k = 1,\ldots, K$. Moreover, it is assumed that the error
term $\varepsilon$ is conditionally independent of $X$, i.e. 
$E( \varepsilon_{g,i} | X_{1,i}, \ldots ,X_{k,i}) = 0$, and that there is
an overlap in observable characteristics across groups ('common support'). 

Estimating the coefficients using OLS, one can fit a counterfactual mean  
$$\overline Y_C = \beta_{0,0} + \sum^K_{k=1}\overline X_{1,k} \widehat \beta_{0,k}$$
that would be observed if group $1$ had the same coefficients like group $0$. By 
adding and subtracting the counterfactual, the observed difference
$$\widehat\Delta^\mu_O = (\overline Y_1 - \overline Y_C) + (\overline Y_C - \overline Y_0) = \widehat\Delta^\mu_S + \widehat\Delta^\mu_C, $$
is divided into the aggregate wage structure effect 
$$\widehat\Delta^\mu_S  = (\widehat \beta_{1,0} - \widehat \beta_{0,0}) + \sum^K_{k=1}\overline X_{1,k}(\widehat \beta_{1,k} - \widehat \beta_{0,k}),$$
that captures outcome differences due to different coefficients, and the composition effect 
$$\widehat\Delta^\mu_X = \sum^K_{k=1} (\overline X_{1,k} - \overline X_{0,k})\widehat \beta_{0,k}, $$
which accounts for mean differences of the covariates. Note
we could also combine coefficients from group 1 with covariates of group 0 to 
define the counterfactual. Such a change in the "reference" group generally
leads to different results.

$\widehat\Delta^\mu_S$ and $\widehat\Delta^\mu_X$ denote the aggregate composition
terms. Since the method hinges on an additive linear model, the terms can be further
decomposed into the contribution of each covariate $k$ in a *detailed
decomposition*. For instance, the contribution of covariate $k = 1$ to the 
structure effect is $\overline X_{1,1}(\widehat \beta_{1,1} - \widehat \beta_{0,1})$
and $(\overline X_{1,1} - \overline X_{0,1})\widehat \beta_{0,1}$ to the composition
effect.

### Reweighting decomposition

The Oaxaca-Blinder decomposition is limited to the mean. Moreover, the 
decomposition terms are biased if the expectation of the outcome conditional
on the covariates is not linear. DiNardo, Fortin, and Lemieux (1996), DFL hereafter,
propose an alternative approach that overcomes both shortcomings. Instead of modeling
the conditional mean, the method uses inverse probability weighting to derive
a counterfactual outcome distribution based on the conditional outcome distribution
of one group and the covariates distribution of the other group. For instance, we 
would reweight the outcome distribution of group 0 such that its covariates
distribution matches that of group 1.
$$F_{Y_C}(y) = \int F_{Y_0}(y|x)dF_{X_1} (x)= \int F_{Y_0}(y|x)\Psi_X(x)dF_{X_0}(x).$$
By applying Bayes' rule, the reweighting factor, 
$$\Psi_X(x) = \frac{dF_{X_1}(x)}{dF_{X_0}(x)} = \frac{P(g=0)P(g=1|x)}{P(g=1)P(g=0|x)},$$
can be rewritten where $P(g)$ and $P(g|x)$ correspond to the (conditional) probabilities of
belonging to group $g$. We can estimate the reweighting factor using sample 
probabilities of each group in the joint sample and conditional probability models (e.g.
logit). The fitted factors are then used to estimate weighted distributional
statistics of interest (e.g., mean, quantiles or Gini coefficient) in 
the reference sample -- group 0 in the present example. The resulting counterfactual
distributional statistic, $\widehat\nu_C=\widehat\nu(F_{Y_C})$, is then again 
contrasted with the observed difference
$$\widehat\Delta_O^{\nu} = (\widehat\nu_1 - \widehat\nu_C) + (\widehat\nu_C - \widehat\nu_0) = \widehat\Delta_S^\nu + \widehat\Delta_X^\nu,$$

which yields again an aggregate wage structure effect and aggregate composition effect.

The two decomposition terms account for the contribution of the covariates and the
conditional outcome distribution, assuming common support and ignorability.
The latter condition asserts that the distribution of unobserved covariates, conditional 
on observed covariates, is independent of the group.

### Sequential decompositions

In contrast to the Oaxaca-Blinder decomposition, where the contributions of each
covariate simply add up, detailed decompositions are not straightforward in the
reweighting framework. However, DFL show that we can sequentially alter the covariates
distributions to decompose the composition effect into the contribution of single
covariates. For instance, assume we want to distinguish the effect of covariate $X_1$ 
(e.g., union status) from that of covariate $X_2$ (e.g., industry).
We begin again with the counterfactual distribution of group 0 with the 
covariates distribution of group 1 
$$F_{Y_{C}}(y) = \iint F_{Y_0}(y|x,z)dF_{X_{1,1}}(x_1|x_2)dF_{X_{1,2}}(x_2)$$

and introduce a second counterfactual where we combine the 
conditional outcome distribution of group 0 as well as the conditional covariate 
distribution of $X_1$ given $X_2$ (e.g., union coverage by industry) of
group 0 with the covariates distribution $X_2$ of group 1
$$F_{Y_{C,X_1}}(y) = \iint F_{Y_0}(y|x,z)dF_{X_{0,1}}(x_1|x_2)dF_{X_{1,2}}(x_2) $$

which can be expressed as the outcome distribution 0
$$F_{Y_{C,X_1}}(y) = \iint F_{Y_0}(y|x_1,x_2)\Psi_{X_1}(x_1|x_2)dF_{X_{0,1}}(x_1|x_2)dF_{X_{0,2}}(x_2),$$
reweighted by the factor
$$\Psi_{X_1}(x_1|x_2) = \frac{dF_{X_{1,1}}(x_1|x_2)}{dF_{X_{0,1}}(x_1|x_2)} =  \frac{P(g=0|x_1)P(g=1|x_1,x_2)}{P(g=1|x_1)P(g=0|x_1,x_2)}.$$

With the distributional statistics of the additional counterfactual, we can 
divide the aggregate decomposition effect into the contribution of each covariate
$$\widehat \Delta_X^{\nu} =  (\widehat \nu_C - \widehat \nu_{C,X_1}) + (\widehat \nu_{C,X_1} - \widehat \nu_0) = \widehat \Delta_{X_1}^\nu + \widehat \Delta_{X_2}^\nu,$$

However, sequential decompositions are path-dependent because the detailed composition
effects attributed to single covariates depend on the order of which we include the 
variables into the sequence. For instance, it matters if we reweight union coverage
by industry ($F_{X_{0,1}} ( x_1 | x_2)$) or the industry given union coverage ($F_{X_{0,2}}( x_2  | x_1)$).
Moreover, we get different results if we derive $\widehat\nu_{C,X_1}$ using the 
conditional covariate distribution from the other group, e.g.
$$F_{Y_{C,X_1}}(y) = \iint F_{Y_0}(y|x_1,x_2)dF_{X_{1,1}}(x_1|x_2)dF_{X_0,2}(x_2).$$

### 'Doubly robust' Oaxca-Blinder

A robust and path-independent alternative...  Barsky et al. 2023

It has also the nice sideeffect to account for potential reweighting errors.




[Barsky et al. (2002)](https://www.jstor.org/stable/3085702) pointed out that OB decompositions do not provide consistent estimates of the wage structure and composition effect if the conditional mean function is non linear. They propose using a reweighting approach as in DiNardo et al. (1996) in the decomposition. In this reweighting approach, the reweighting function $\widehat\Psi(X_i)$ makes the characteristics of group $A$ similar to those of group $B$, defined as $X^C_A$. In addition, the OLS regression coefficients are estimated for the reweighted group $\overline X^C_A$, yielding $\widehat \beta^C_A$. Analogous, group A can be used as reference group, thereby reweighting the characteristics of group $B$ would be reweighted to reflect the characteristics of group $A$. 

To decompose the overall group difference in the reweighted regression decomposition, $\overline X^C_A\widehat \beta^C_A$ is used as counterfactual scenario. The overall gap is then decomposed as: 

$$
\begin{aligned} \widehat\Delta^\mu_{O,R} &= (\widehat \beta_{B0} - \widehat \beta^C_{A0}) + \sum^K_{k=1} (\overline X_{Bk}\widehat \beta_{Bk} - \overline X^C_{Ak}\widehat \beta^C_{Ak})) + \sum^K_{k=1} (\overline X^C_{Ak}\beta^C_{Ak} - \overline X_{Ak}\widehat \beta_{Ak}) \\
 &= \widehat\Delta^\mu_{S,R}  + \widehat\Delta^\mu_{X,R}. 
\end{aligned}
$$
These decompostion parts are further divided into pure structure and composition effects, $\widehat\Delta^\mu_{S,p}$ and $\widehat\Delta^\mu_{X,p}$ respectively, and into reweighting error $\widehat\Delta^\mu_{S,e}$ and specification error $\widehat\Delta^\mu_{X,e}$. The structure effect $\widehat\Delta^\mu_{S,R}$ can be written as 

$$\widehat\Delta^\mu_{S,R} = \underbrace{(\widehat \beta_{B0} - \widehat \beta^C_{A0}) + \sum^K_{k=1}\overline X_{Bk}(\widehat \beta_{Bk} - \widehat \beta^C_{Ak})}_{\widehat\Delta^\mu_{S,p} \text{ (Pure structure effect)}} + \underbrace{\sum^K_{k=1} (\overline X_{Bk} - \overline X^C_{Ak})\widehat \beta^C_{Ak}.}_{\widehat\Delta^\mu_{S,e} \text{ (Reweighting error)}}$$

Similarly, the composition effect can be written as

$$\widehat\Delta^\mu_{X,R} = \underbrace{\sum^K_{k=1} (\overline X^C_{Ak} - \overline X_{Ak})\widehat \beta_{Ak}}_{\widehat\Delta^\mu_{X,p} \text{ (Pure composition effect)}} + \underbrace{(\widehat \beta^C_{A0} - \widehat \beta_{A0}) + \sum^K_{k=1}\overline X^C_{Ak}(\widehat \beta^C_{Ak} - \widehat \beta_{Ak}).}_{\widehat\Delta^\mu_{X,e} \text{ (Specification error)}}$$

The reweighted regression decomposition enhances the original OB decomposition method in two aspects. First, the structure effect is estimated by comparing $\widehat \beta_{B0}$ with the reweighted estimates of group $A$, $\widehat \beta^C_{A0}$ instead of $\widehat \beta_{A0}$. Therefore, the pure wage structure effect entails the true underlying structural differences in group $A$ and $B$ (Fortin et al. 2011). The reweighting error indicates how accurate the reweighting function $\widehat\Psi(X_i)$ and will be zero if the composition of group $B$ is equal to the reweighted group $A$.  

Second, the pure composition effect displays the differences between the composition of group $A$ and the reweighted group $A$. If the model is truly linear and correctly specified, the specification error is zero and $\widehat\Delta^\mu_{X,p} = \widehat\Delta^\mu_{X,R}$. If the specification error is not close to zero, it is advisable to revise the model specification. While the original OB decomposition only allows for decomposition at the mean, the following methods provide means to decompose other distributional metrics. 



### RIF decomposition 

Path independent and going beyond the mean


To estimate decompositions beyond the mean, we implemented a reweighted RIF regression decomposition as proposed by Firpo et al. (2018). Firpo et al. propose to approximate the conditional expectation of the RIF given the explanatory variables with a linear regression. The regression coefficients can be consistent estimates of the average derivatives $\widehat{\alpha}(\nu)$ if the conditional expectations of the RIF are linear in $X$ (see [Firpo et al., 2009a](https://www.econometricsociety.org/publications/econometrica/2009/05/01/unconditional-quantile-regressions), [Rothe 2015: 328](https://doi.org/10.1080/07350015.2014.948959)).

In the reweighted RIF regression decompositions, first the RIF of the outcome variable $Y$ and a distributional statistic of interest $\nu$ is computed. Then, an OLS regression of the transformed outcome variable $Y$ is run on the explanatory variables $X$. Thereafter, the decompostion method is analogous to the original OB method:

$$\widehat\Delta^\nu_{O,R} = \underbrace{(\widehat \beta_{B0} - \widehat \beta^C_{A0}) + \sum^K_{k=1}\overline X_{Bk}(\widehat \beta_{Bk} - \widehat \beta^C_{Ak})}_{\widehat\Delta^\nu_{S,p} \text{ (Pure structure effect)}} + \underbrace{\sum^K_{k=1} (\overline X_{Bk} - \overline X^C_{Ak})\widehat \beta^C_{Ak}}_{\widehat\Delta^\nu_{S,e} \text{ (Reweighting error)}} + \underbrace{\sum^K_{k=1} (\overline X^C_{Ak} - \overline X_{Ak})\widehat \beta_{Ak}}_{\widehat\Delta^\nu_{X,p} \text{ (Pure composition effect)}} + \underbrace{(\widehat \beta^C_{A0} - \widehat \beta_{A0}) + \sum^K_{k=1}\overline X^C_{Ak}(\widehat \beta^C_{Ak} - \widehat \beta_{Ak}).}_{\widehat\Delta^\nu_{X,e} \text{ (Specification error)}}$$
with $\widehat \beta$ being the coefficients of the OLS regression of the transformed outcome variable $Y$ run on the explanatory variables $X$ of group $A$ and $B$. 


By default, the RIF of quantiles, the mean, the variance, the Gini coefficient, the interquantile range and the quantile ratio are available in `ddeco`. Moreover, the package allows to calculate the RIF for additional statistics with user-written functions (see example below). [Cowell and Flachaire (2007)](https://doi.org/10.1016/j.jeconom.2007.01.001), [Essama-Nssah & Lambert (2012)](https://doi.org/10.1108/S1049-2585(2012)0000020009), and [Rios-Avila (2020)](https://doi.org/10.1177/1536867X20909690) derive the influence functions for an array of distributional statistics. More further information on RIF regressions can be found in the documentation of the [`rifreg`](https://github.com/samumei/rifreg) package written by the same authors as `ddeco`.

It is possible to compute the RIF regression decomposition without reweighting. However, it is strongly advisable to use reweighting and assess the reweighting and specification error, as large errors highlight inconsistent estimations.


### Inference

`ddeco` allows to bootstrap standard errors in both the `ob_deco()` and `dfl_deco()` function. Analytical standard errors can be nontrivial when the RIF introduces an additional estimation step. In particular, this is the case for quantiles where the density has to be estimated (see [Firpo, Fortin, and Lemieux, 2009b](https://www.econometricsociety.org/publications/econometrica/2009/05/01/unconditional-quantile-regressions/supp/6822_extensions_0.pdf)). 





# Old




### Example

Fortin, Lemieux, and Firpo (FLF, 2011, p. 79-88) decompose the increase in US 
male wage inequality between the early 1980s and the early 2000s using the CPS
data. In this example, we replicate their results and treat the observations from
1983 to 1985 as reference group that is reweighted.

```{r, eval = FALSE}
library(ddeco)
data("men8305")
flf_model <- log(wage) ~ union*(education + experience) + education*experience
flf_male_inequality  <- dfl_deco(flf_model,
                                  data = men8305,
                                  weights = weights,
                                  group = year)
```
We can summarize the results:
```{r, eval = FALSE}
summary(flf_male_inequality)
```

Using `plot()`, we can illustrate the decomposition accross different quantiles.
```{r, eval = FALSE}
plot(flf_male_inequality)
```



## Examples

The following examples illustrate the workings of the main decomposition functions in `ddeco`. We use a sample of the National Longitudinal Survey (NLSY) 79 containig wage data from 2000 of workers who were aged 35 to 43 in that year. The data is from O'Neill and O'Neill (2006) and is used as an illustration of the Oxaca-Blinder mean decomposition in Fortin, Lemieux, and Firpo (2011). The data contains 2655 male and 2654 female observations, respectively.

```{r, eval = FALSE} 
library(ddeco)
data("nlys00")
```


### Oaxaca-Blinder Decomposition 

We are interested in the change of mean log hourly wages from 1988-1990 and 2014-2016. The change is decomposed into a composition and wage structure effect. 

The parameter `statistic` specifies quantiles as our distributional statistic of interest, while `probs` defines the probabilities of the quantiles.

```{r, eval = FALSE}
  

  model <- log(wage) ~ age + region + black + hispanic + education + years_worked_civilian + part_time + industry

  ob_deco_estimate <- ob_deco(formula = model,
                                      data = nlys00,
                                      group = female)
```

The `summary` method returns the estimated coefficients and the standard errors. Per default, `summary` returns heteroscedasticity-consistent standard errors estimated if the variance is not bootstrapped.
```{r, eval=FALSE}
summary(ob_deco_estimate)
```

`ddeco` comes with a convenient plot function. It illustrates the estimated UQPE across the distribution. The confidence bands are based on the same standard errors as returned by `summary`. 

-> DO WE NEED PLOT FUNCTION FOR CLASSIC OB DECO?
```{r, eval=FALSE}
plot(ob_deco_estimate, varselect = c("education"))
```


#### Maybe add Reweighting, Weights, and Changing the Reference Group examples

To add reweighting in 
Setting `bootstrap=TRUE` bootstraps standard errors by resampling from all observations and reestimating both the RIF and the regression in every iteration. We can set number of `bootstrap_iterations` and the number of `cores`.  

```{r, eval=FALSE}
  model <- log(wage) ~ age + region + black + hispanic + education + years_worked_civilian + part_time + industry

  ob_deco_estimate <- ob_deco(formula = model,
                              data = nlys00,
                              group = female, 
                              reweighting = TRUE)


```



### Reweighted RIF Regression Decomposition

The same function is used to decompose RIF regressions. Setting `rifreg = TRUE` and indicating `rifreg_statistic`, decomposes the gap at the specified statistic. The default is `rifreg_statistic = "quantiles"` and `rifreg_probs = c(1:9)/10`, meaning the gap is decomposed at each decile. 

```{r, eval=FALSE}
  model <- log(wage) ~ age + region + black + hispanic + education + years_worked_civilian + part_time + industry

  ob_deco_estimate <- ob_deco(formula = model,
                              data = nlys00,
                              group = female, 
                              reweighting = TRUE, 
                              rifreg = TRUE,
                              bootstrap = TRUE,
                              bootstrap_iteration = 10)
```





To decompose the the difference in the Gini coefficient of two groups, `rifreg_statistic = "gini"` needs to be indicated accordingly. The RIF functions for the following functions are currently implemented: `"quantiles"`, `"mean"`, `"variance"`, `"gini"`, `"interquantile_range"`, and `"interquantile_ratio"`. However, `ob_deco` also allows to pass a custom RIF function for the decomposition, by setting `rifreg_statistic = "custom"` and passing the custom function to `custom_rif_function`. 


## Credits

David Gallusser & Samuel Meier

## References

Barsky, Robert, John Bound, Kerwin Kofi Charles, and Joseph P. Lupton. 2002. “Accounting for the Black-White Wealth Gap: A Nonparametric Approach.” *Journal of the American Statistical Association* 97: 663–73.

Cowell, Frank A., and Emmanuel Flachaire. 2007. "Income distribution and inequality measurement: The problem of extreme values." *Journal of Econometrics* 141: 1044–1072.

Essama-Nssah, Boniface, and Peter J. Lambert. 2012. “Influence Functions for Policy Impact Analysis.” In John A. Bishop and Rafael Salas, eds., *Inequality, Mobility and Segregation: Essays in Honor of Jacques Silber*. Bigley, UK: Emerald Group Publishing
Limited.

Firpo, Sergio, Nicole M. Fortin, and Thomas Lemieux. 2007a. “Unconditional Quantile Regressions.” *Technical Working Paper 339, National Bureau of Economic Research*. Cambridge, MA. 

Firpo, Sergio, Nicole M. Fortin, and Thomas Lemieux. 2009a. "Unconditional Quantile Regressions." *Econometrica* 77(3): 953-973.

Firpo, Sergio, Nicole M. Fortin, and Thomas Lemieux. 2009b. "Supplement to 'Unconditional Quantile Regressions'." *Econometrica Supplemental Material*, 77.

Fortin, Nicole M., Thomas Lemieux, and Sergio Firpo. 2011. "Decomposition Methods in Economics." *National Bureau of Economic Research - Working Paper Series*, 16045.

Rios-Avila, Fernando (2020): "Recentered influence functions (RIFs) in Stata: RIF regression and RIF decomposition." *The Stata Journal* 20(1): 51-94.

Rothe, Christoph (2015): "Decomposing the Composition Effect. The Role of Covariates in Determining Between-Group Differences in Economic Outcomes." *Journal of Business & Economic Statistics* 33(3): 323-337.


