---
output:
  github_document:
    toc: true
    toc_depth: 3
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# `ddeco`: Detailed decompositions of between-group differences in R

## Overview

**ddeco** implements the [Oaxaca (1973)](https://www.jstor.org/stable/2525981)-[Blinder (1973)](https://www.jstor.org/stable/144855) decomposition method and generalizations
of it that decompose differences in distributional statistics beyond the mean.

`ob_deco()` divides differences in the mean outcome between two groups into one 
part explained by different covariates (composition effect) and into 
another part due to differences in the way covariates are linked to the 
outcome variable (wage structure effect). The function further divides the two
effects into the contribution of each covariate and allows for weighted 'doubly
robust' decompositions. For distributional statistics beyond the mean,
the function performs the RIF decomposition proposed by [Firpo, Fortin, and Lemieux (2018)]( https://doi.org/10.3390/econometrics6020028). 

`dfl_deco()` divides differences in distributional statistics into an
composition effect and a wage structure effect using inverse probability
weighting as introduced by  [DiNardo, Fortin, and Lemieux (1996)](https://www.jstor.org/stable/2171954). The function also allows
to sequentially decompose the composition effect into the contribution of
single covariates.

The package contains generic summary, print and plot functions for the results
and computes bootstrapped standard errors. This documentation provides a brief 
overview over the functions implemented in the package. For a more detailed 
discussion of decomposition methods including their respective assumptions and 
limitations, refer to [Fortin, Lemieux, and Firpo (2011)](https://economics.ubc.ca/wp-content/uploads/sites/38/2013/05/pdf_paper_nicole-fortin-decomposition-methods.pdf) and Firpo et al. (2018). 

## Installation 

You can either install the CRAN version of `ddeco`
```{r, eval=FALSE}
install.packages("ddeco")
```

or the latest development version from GitHub:
```{r , eval=FALSE}
devtools::install_github("samumei/ddeco")
```

## Background

### Oaxaca-Blinder Decomposition (and Reweighted Variant)

<span style="color:red">
The original decomposition method introduced by Oaxaca (1973) and Blinder (1973) 
decomposes the difference in the mean of an outcome variable (e.g. wage) between
two groups $g = A, B$ into a part explained by differences in covariate means (e.g. 
educational level or experience) and another part due different linear
regression coefficients (e.g. returns to education) linking covariates to the
outcome. 
</span>


<span style="color:red">
The method linearly models the relationship between the outcome $Y$ and covariates
$X$, i.e. 
$$Y_{g,i} = \beta_{g,0} + \sum^K_{k=1}X_{k,i}\beta_{g,k} + \varepsilon_{g,i},$$
where $\beta_{g,0}$ and $\beta_{g,k}$ and the intercept and slope coefficents of
covariates $k = 1,..., K$, respectively. Moreover, it is assumed that the error
term $\varepsilon$ is conditionally independent of $X$, i.e. 
$E( \varepsilon_{g,i} | X_{1,i} \ldots X_{k,i}) = 0$, and that there is
an overlap in observable characteristics across groups ('common support'). 
</span>


<span style="color:red">
Estimating the coefficients with OLS, one can fit a counterfactual mean  
$$\overline Y_C = \beta_{A,0} + \sum^K_{k=1}\overline X_{B,k} \widehat \beta_{A,k}$$
that would be observed if group $B$ had the same coefficients as group $A$. By 
adding and substracting the counterfactual, the observed difference is divided 
$$\widehat\Delta^\mu_O = (\overline Y_B - \overline Y_C) + (\overline Y_C - \overline Y_A) = \widehat\Delta^\mu_S + \widehat\Delta^\mu_C, $$
into the aggregate wage structure effect 
$$\widehat\Delta^\mu_S  = (\widehat \beta_{B0} - \widehat \beta_{A0}) + \sum^K_{k=1}\overline X_{Bk}(\widehat \beta_{Bk} - \widehat \beta_{Ak}),$$
that captures outcome differences due to differences in coefficients, and the composition effect 
$$\widehat\Delta^\mu_X = \sum^K_{k=1} (\overline X_{Bk} - \overline X_{Ak})\widehat \beta_{Ak}. $$
which identifies mean differencs attributed to different covariate means. Note
we could also combine coefficients from group B with covariates of group A to 
perform the decomposition, instead. Such a change in the 'reference' group generally
leads to different results.
</span>


The original decomposition method introduced by Oaxaca (1973) and Blinder (1973) decomposes group differences in the mean of an outcome variable (e.g. wage). The group differences in the mean $\widehat\Delta^\mu_O$ are decomposed into two parts. One part, the "composition effect" $\widehat\Delta^\mu_X$, is explained by differences in the observable covariates (e.g. educational level or experience). The other part, the "structure effect" $\widehat\Delta^\mu_S$, is due to different returns to these covariates. 

The observed difference can be written as
$$\widehat\Delta^\mu_O = \overline Y_B - \overline Y_A,$$

with the composition and structure effect as:

$$\widehat\Delta^\mu_O = \underbrace{(\widehat \beta_{B0} - \widehat \beta_{A0}) + \sum^K_{k=1}\overline X_{Bk}(\widehat \beta_{Bk} - \widehat \beta_{Ak})}_{\widehat\Delta^\mu_S \text{ (Unexplained)}} + \underbrace{\sum^K_{k=1} (\overline X_{Bk} - \overline X_{Ak})\widehat \beta_{Ak}}_{\widehat\Delta^\mu_X \text{ (Explained)}} $$
where $\beta_{g}$ are the intercept and slope coefficents of covariates $k = 1,..., K$, in the OLS regression models of group $g = A, B$. In this notation. The counterfactual reference scenario is $\overline X_{Bk}\beta_{Ak}$, the covariates' mean of group $B$, mutliplied by the covariates' coefficients of group $A$. However, the reference group could also be the set as $\overline X_{Ak}\beta_{Bk}$.

The terms $\widehat\Delta^\mu_S $ and $\widehat\Delta^\mu_X$ denote the \emph{aggregate decomposition}, i.e. the overall structure and composition effect. In the \emph{detailed decomposition}, the contribution of each covariate $k$ to the structure and composition effect is estimated. For instance, the contribution of covariate $k = 1$ to the structure effect is $\overline X_{B1}(\widehat \beta_{B1} - \widehat \beta_{A1})$ and $(\overline X_{B1} - \overline X_{A1})\widehat \beta_{A1}$ to the composition effect.




## Background (By DG)

The **ddeco** package allows to decompose observed differences in a distributional
statistics $\nu_t=\nu(F_{Y_t})$ of an outcome variable $Y_t$ (e.g., hourly wages) between 
two groups $t = 0,1$.
$$\Delta_O^{\nu} = \nu_1 - \nu_0.$$
We are interested to what extent the difference can explained by a different composition
of observable covariates and by differences in structure linking covariates to the 
outcome variables. For this purpose, we define a conterfactual outcome statistic that
would be observed if group 0 had the same composition like group 1 but otherwise
the same structure linking covariates to the outcome variable. By contrasting 
the observed statistics to the counterfactual, we can indentify the composition 
effect $\Delta_S^\nu$ and the (wage) structure effect $\Delta_X^\nu$  that capture 
differencs in the structure linking covariates to the outcome and in the difference
in the composition of covariates, respectively: 
$$\Delta_O^{\nu} = (\nu_1 - \nu_C) + (\nu_C - \nu_0) = \Delta_S^\nu + \Delta_X^\nu.$$
Note, we could specifiy the counterfactual the other way round.

Under certain assumptions (see Fortin, Lemieux, and Firpo 2011)
(ignorability/CIA, overlapping support), we can derive...

Standard OB case.
...



#### Reweighted Regression Decompostion

[Barsky et al. (2002)](https://www.jstor.org/stable/3085702) pointed out that OB decompositions do not provide consistent estimates of the wage structure and composition effect if the conditional mean function is non linear. They propose using a reweighting approach as in DiNardo et al. (1996) in the decomposition. In this reweighting approach, the reweighting function $\widehat\Psi(X_i)$ makes the characteristics of group $A$ similar to those of group $B$, defined as $X^C_A$. In addition, the OLS regression coefficients are estimated for the reweighted group $\overline X^C_A$, yielding $\widehat \beta^C_A$. Analogous, group A can be used as reference group, thereby reweighting the characteristics of group $B$ would be reweighted to reflect the characteristics of group $A$. 

To decompose the overall group difference in the reweighted regression decomposition, $\overline X^C_A\widehat \beta^C_A$ is used as counterfactual scenario. The overall gap is then decomposed as: 

$$
\begin{aligned} \widehat\Delta^\mu_{O,R} &= (\widehat \beta_{B0} - \widehat \beta^C_{A0}) + \sum^K_{k=1} (\overline X_{Bk}\widehat \beta_{Bk} - \overline X^C_{Ak}\widehat \beta^C_{Ak})) + \sum^K_{k=1} (\overline X^C_{Ak}\beta^C_{Ak} - \overline X_{Ak}\widehat \beta_{Ak}) \\
 &= \widehat\Delta^\mu_{S,R}  + \widehat\Delta^\mu_{X,R}. 
\end{aligned}
$$
These decompostion parts are further divided into pure structure and composition effects, $\widehat\Delta^\mu_{S,p}$ and $\widehat\Delta^\mu_{X,p}$ respectively, and into reweighting error $\widehat\Delta^\mu_{S,e}$ and specification error $\widehat\Delta^\mu_{X,e}$. The structure effect $\widehat\Delta^\mu_{S,R}$ can be written as 

$$\widehat\Delta^\mu_{S,R} = \underbrace{(\widehat \beta_{B0} - \widehat \beta^C_{A0}) + \sum^K_{k=1}\overline X_{Bk}(\widehat \beta_{Bk} - \widehat \beta^C_{Ak})}_{\widehat\Delta^\mu_{S,p} \text{ (Pure structure effect)}} + \underbrace{\sum^K_{k=1} (\overline X_{Bk} - \overline X^C_{Ak})\widehat \beta^C_{Ak}.}_{\widehat\Delta^\mu_{S,e} \text{ (Reweighting error)}}$$

Similarly, the composition effect can be written as

$$\widehat\Delta^\mu_{X,R} = \underbrace{\sum^K_{k=1} (\overline X^C_{Ak} - \overline X_{Ak})\widehat \beta_{Ak}}_{\widehat\Delta^\mu_{X,p} \text{ (Pure composition effect)}} + \underbrace{(\widehat \beta^C_{A0} - \widehat \beta_{A0}) + \sum^K_{k=1}\overline X^C_{Ak}(\widehat \beta^C_{Ak} - \widehat \beta_{Ak}).}_{\widehat\Delta^\mu_{X,e} \text{ (Specification error)}}$$

The reweighted regression decomposition enhances the original OB decomposition method in two aspects. First, the structure effect is estimated by comparing $\widehat \beta_{B0}$ with the reweighted estimates of group $A$, $\widehat \beta^C_{A0}$ instead of $\widehat \beta_{A0}$. Therefore, the pure wage structure effect entails the true underlying structural differences in group $A$ and $B$ (Fortin et al. 2011). The reweighting error indicates how accurate the reweighting function $\widehat\Psi(X_i)$ and will be zero if the composition of group $B$ is equal to the reweighted group $A$.  

Second, the pure composition effect displays the differences between the composition of group $A$ and the reweighted group $A$. If the model is truly linear and correctly specified, the specification error is zero and $\widehat\Delta^\mu_{X,p} = \widehat\Delta^\mu_{X,R}$. If the specification error is not close to zero, it is advisable to revise the model specification. While the original OB decomposition only allows for decomposition at the mean, the following methods provide means to decompose other distributional metrics. 



### Reweighted RIF Regression Decomposition 

To estimate decompositions beyond the mean, we implemented a reweighted RIF regression decomposition as proposed by Firpo et al. (2018). Firpo et al. propose to approximate the conditional expectation of the RIF given the explanatory variables with a linear regression. The regression coefficients can be consistent estimates of the average derivatives $\widehat{\alpha}(\nu)$ if the conditional expectations of the RIF are linear in $X$ (see [Firpo et al., 2009a](https://www.econometricsociety.org/publications/econometrica/2009/05/01/unconditional-quantile-regressions), [Rothe 2015: 328](https://doi.org/10.1080/07350015.2014.948959)).

In the reweighted RIF regression decompositions, first the RIF of the outcome variable $Y$ and a distributional statistic of interest $\nu$ is computed. Then, an OLS regression of the transformed outcome variable $Y$ is run on the explanatory variables $X$. Thereafter, the decompostion method is analogous to the original OB method:

$$\widehat\Delta^\nu_{O,R} = \underbrace{(\widehat \beta_{B0} - \widehat \beta^C_{A0}) + \sum^K_{k=1}\overline X_{Bk}(\widehat \beta_{Bk} - \widehat \beta^C_{Ak})}_{\widehat\Delta^\nu_{S,p} \text{ (Pure structure effect)}} + \underbrace{\sum^K_{k=1} (\overline X_{Bk} - \overline X^C_{Ak})\widehat \beta^C_{Ak}}_{\widehat\Delta^\nu_{S,e} \text{ (Reweighting error)}} + \underbrace{\sum^K_{k=1} (\overline X^C_{Ak} - \overline X_{Ak})\widehat \beta_{Ak}}_{\widehat\Delta^\nu_{X,p} \text{ (Pure composition effect)}} + \underbrace{(\widehat \beta^C_{A0} - \widehat \beta_{A0}) + \sum^K_{k=1}\overline X^C_{Ak}(\widehat \beta^C_{Ak} - \widehat \beta_{Ak}).}_{\widehat\Delta^\nu_{X,e} \text{ (Specification error)}}$$
with $\widehat \beta$ being the coefficients of the OLS regression of the transformed outcome variable $Y$ run on the explanatory variables $X$ of group $A$ and $B$. 


By default, the RIF of quantiles, the mean, the variance, the Gini coefficient, the interquantile range and the quantile ratio are available in `ddeco`. Moreover, the package allows to calculate the RIF for additional statistics with user-written functions (see example below). [Cowell and Flachaire (2007)](https://doi.org/10.1016/j.jeconom.2007.01.001), [Essama-Nssah & Lambert (2012)](https://doi.org/10.1108/S1049-2585(2012)0000020009), and [Rios-Avila (2020)](https://doi.org/10.1177/1536867X20909690) derive the influence functions for an array of distributional statistics. More further information on RIF regressions can be found in the documentation of the [`rifreg`](https://github.com/samumei/rifreg) package written by the same authors as `ddeco`.

It is possible to compute the RIF regression decomposition without reweighting. However, it is strongly advisable to use reweighting and assess the reweighting and specification error, as large errors highlight inconsistent estimations.



### Decomposition Through Reweighting (DiNardo, Fortin, and Lemieux 1996)

Another method to decompose group differences in an outcome variable beyond the mean is using inverse propensity reweighting as proposed by DiNardo, Fortin, and Lemieux (1996). We implement this decompostion method in the function `dfl_deco()`. The procedure reweights the sample distribution of a reference group such that the group's covariates distribution matches the covariates distribution of a counterfactual group. Reweighting factors are derived by modelling the probability of belonging to the one group instead of the other conditional on covariates. The function allows detailed decompositions of the composition effect by sequentially reweighting (conditional) covariate distributions. 

If group $A$ is set as reference, it is reweighted to match the distribution of characteristics of group $B$. First, a logit model is run to estimate the probability of belonging to group $B$ conditional on covariates: 

$$ Pr(D_B = 1 | X) = 1 - Pr(D_A = 1 | X ). $$

Next, the reweighting factor $\widehat\Psi(X)$ for observations in group $A$ is estimated:

$$ \widehat\Psi(X_i) = \frac{\widehat Pr(D_B = 1 | X) / \widehat Pr(D_B = 1) }{\widehat Pr(D_B = 0 | X) / \widehat Pr(D_B = 0)}.$$
Note that the same estimation method for $\widehat\Psi(X)$ is applied when computing the reweighting factor for the reweighted OB and RIF regression decomposition discussed above. Using $\widehat\Psi(X)$ to reweight the sample obersavtions in group $A$, the counterfactual statistic of interest then be computed. For instance, to estimate the overall wage gap at a specific quantile $Q_X$ and decompose it into aggregate structure and composition effect following approach is applied: 

$$
\begin{aligned}
\widehat\Delta^{Q_X}_O &= Q_{B,X} - Q_{A,X} \\
&= \underbrace{Q^C_{A,X}- Q_{A,X} }_{\widehat\Delta^{Q_X}_S \text{ (Unexplained)}} + \underbrace{Q_{B,X} - Q^C_{A,X}}_{\widehat\Delta^{Q_X}_X \text{ (Explained)}}
\end{aligned}
$$

with $Q_{gX}$ being the estimated Quantile in groups $g = A, B$ and $Q^C_AX$ being the quantile estimate in the reweighted group $A$. For further details on this method, including the detailed decomposition, refer to DiNardo et al. (1996) and Fortin et al. (2011, p. 63-69). 



## Reweighting decompositions with `dfl_deco()`

### Aggregate decomposition

`dfl_deco()` uses inverse probability weighting to estimate counterfactual 
distributional statistics. The counterfactual combines the conditional outcome
distribution of the reference group, here defined as group 0, with the covariates 
distribution of the comparison group, and can expressed as reweighted version of the
reference distribution
$$F_{Y_C}(y) = \int F_{Y_0}(y|x)dF_{X_1} (x)= \int F_{Y_0}(y|x)\Psi(x)dF_{X_0}(x),$$

where the reweighting factor $\Psi(x) = \frac{P(t=0)P(t=1|X)}{P(t=1)P(t=0|X)}$ 
captures the inverse probability of belonging to group 1. The reweighting factor
can be written as function of the probabilities of the binary group 
variable $t$ and can, thus, readily estimated in the pooled sample of the two
group unsing conditional probability models. The counterfactual statistics are
then estimated with weighted estimator using the observed data of group 0 and 
the fitted reweighting factors.

### Sequential decomposition

We can extend the the reweighting approach to sequentially decompose the composition
effect into the contribution of single covariates $X$ and $Z$, respectively 
$$\Delta_X^{\nu} =  (\nu_C - \nu_{C,X}) + (\nu_{C,X} - \nu_0) = \Delta_{X,X}^\nu + \Delta_{X,Z}^\nu,$$

where we define $\nu_C$ as the statistic of the counterfactual where we combine the 
conditional outcome of group 0 with the covariates distribution of group 1, i.e.
$$F_{Y_{C}}(y) = \iint F_{Y_0}(y|x,z)dF_{X_1}(x|z)dF_{Z_1}(x).$$

For the decomposition work, we require an additional counterfactual $\nu_{C,X}$ that 
combines the conditional outcome distribution of group 0 and the distribution of covariate $X$ given $Z$ of the same group with the marginal distribution of $Z$ of group 1, i.e.
$$F_{Y_{C,X}}(y) = \iint F_{Y_0}(y|x,z)dF_{X_0}(x|z)dF_{Z_1}(x).$$
Note, sequential decompositions are path-dependent, i.e. the detailed composition
effects attributed to single covariates depend on the order we include the 
variables into the sequence (i.e. if we integrate over the conditional distribution
of $X$ or that of $Z$, respecitvely). Moreover, we get different results if we 
derive $\nu_{C,X}$ using the conditional covariate distribution from the other group, e.g.
$$F_{Y_{C,X}}(y) = \iint F_{Y_0}(y|x,z)dF_{X_1}(x|z)dF_{Z_0}(x).$$


### Example

Fortin, Lemieux, and Firpo (FLF, 2011, p. 79-88) decompose the increase in US 
male wage inequality between the early 1980s and the early 2000s using the CPS
data. In this example, we replicate their results and treat the observations from
1983 to 1985 as reference group that is reweighted.

```{r, eval = FALSE}
library(ddeco)
data("men8305")
flf_model <- log(wage) ~ union*(education + experience) + education*experience
flf_male_inequality  <- dfl_deco(flf_model,
                                  data = men8305,
                                  weights = weights,
                                  group = year)
```
We can summarize the results:
```{r, eval = FALSE}
summary(flf_male_inequality)
```

Using `plot()`, we can illustrate the decomposition accross different quantiles.
```{r, eval = FALSE}
plot(flf_male_inequality)
```


### Inference

`ddeco` allows to bootstrap standard errors in both the `ob_deco()` and `dfl_deco()` function. Analytical standard errors can be nontrivial when the RIF introduces an additional estimation step. In particular, this is the case for quantiles where the density has to be estimated (see [Firpo, Fortin, and Lemieux, 2009b](https://www.econometricsociety.org/publications/econometrica/2009/05/01/unconditional-quantile-regressions/supp/6822_extensions_0.pdf)). 


## Examples

The following examples illustrate the workings of the main decomposition functions in `ddeco`. We use a sample of the National Longitudinal Survey (NLSY) 79 containig wage data from 2000 of workers who were aged 35 to 43 in that year. The data is from O'Neill and O'Neill (2006) and is used as an illustration of the Oxaca-Blinder mean decomposition in Fortin, Lemieux, and Firpo (2011). The data contains 2655 male and 2654 female observations, respectively.

```{r, eval = FALSE} 
library(ddeco)
data("nlys00")
```


### Oaxaca-Blinder Decomposition 

We are interested in the change of mean log hourly wages from 1988-1990 and 2014-2016. The change is decomposed into a composition and wage structure effect. 

The parameter `statistic` specifies quantiles as our distributional statistic of interest, while `probs` defines the probabilities of the quantiles.

```{r, eval = FALSE}
  

  model <- log(wage) ~ age + region + black + hispanic + education + years_worked_civilian + part_time + industry

  ob_deco_estimate <- ob_deco(formula = model,
                                      data = nlys00,
                                      group = female)
```

The `summary` method returns the estimated coefficients and the standard errors. Per default, `summary` returns heteroscedasticity-consistent standard errors estimated if the variance is not bootstrapped.
```{r, eval=FALSE}
summary(ob_deco_estimate)
```

`ddeco` comes with a convenient plot function. It illustrates the estimated UQPE across the distribution. The confidence bands are based on the same standard errors as returned by `summary`. 

-> DO WE NEED PLOT FUNCTION FOR CLASSIC OB DECO?
```{r, eval=FALSE}
plot(ob_deco_estimate, varselect = c("education"))
```


#### Maybe add Reweighting, Weights, and Changing the Reference Group examples

To add reweighting in 
Setting `bootstrap=TRUE` bootstraps standard errors by resampling from all observations and reestimating both the RIF and the regression in every iteration. We can set number of `bootstrap_iterations` and the number of `cores`.  

```{r, eval=FALSE}
  model <- log(wage) ~ age + region + black + hispanic + education + years_worked_civilian + part_time + industry

  ob_deco_estimate <- ob_deco(formula = model,
                              data = nlys00,
                              group = female, 
                              reweighting = TRUE)


```



### Reweighted RIF Regression Decomposition

The same function is used to decompose RIF regressions. Setting `rifreg = TRUE` and indicating `rifreg_statistic`, decomposes the gap at the specified statistic. The default is `rifreg_statistic = "quantiles"` and `rifreg_probs = c(1:9)/10`, meaning the gap is decomposed at each decile. 

```{r, eval=FALSE}
  model <- log(wage) ~ age + region + black + hispanic + education + years_worked_civilian + part_time + industry

  ob_deco_estimate <- ob_deco(formula = model,
                              data = nlys00,
                              group = female, 
                              reweighting = TRUE, 
                              rifreg = TRUE,
                              bootstrap = TRUE,
                              bootstrap_iteration = 10)
```





To decompose the the difference in the Gini coefficient of two groups, `rifreg_statistic = "gini"` needs to be indicated accordingly. The RIF functions for the following functions are currently implemented: `"quantiles"`, `"mean"`, `"variance"`, `"gini"`, `"interquantile_range"`, and `"interquantile_ratio"`. However, `ob_deco` also allows to pass a custom RIF function for the decomposition, by setting `rifreg_statistic = "custom"` and passing the custom function to `custom_rif_function`. 


## Credits

David Gallusser & Samuel Meier

## References

Barsky, Robert, John Bound, Kerwin Kofi Charles, and Joseph P. Lupton. 2002. “Accounting for the Black-White Wealth Gap: A Nonparametric Approach.” *Journal of the American Statistical Association* 97: 663–73.

Cowell, Frank A., and Emmanuel Flachaire. 2007. "Income distribution and inequality measurement: The problem of extreme values." *Journal of Econometrics* 141: 1044–1072.

Essama-Nssah, Boniface, and Peter J. Lambert. 2012. “Influence Functions for Policy Impact Analysis.” In John A. Bishop and Rafael Salas, eds., *Inequality, Mobility and Segregation: Essays in Honor of Jacques Silber*. Bigley, UK: Emerald Group Publishing
Limited.

Firpo, Sergio, Nicole M. Fortin, and Thomas Lemieux. 2007a. “Unconditional Quantile Regressions.” *Technical Working Paper 339, National Bureau of Economic Research*. Cambridge, MA. 

Firpo, Sergio, Nicole M. Fortin, and Thomas Lemieux. 2009a. "Unconditional Quantile Regressions." *Econometrica* 77(3): 953-973.

Firpo, Sergio, Nicole M. Fortin, and Thomas Lemieux. 2009b. "Supplement to 'Unconditional Quantile Regressions'." *Econometrica Supplemental Material*, 77.

Fortin, Nicole M., Thomas Lemieux, and Sergio Firpo. 2011. "Decomposition Methods in Economics." *National Bureau of Economic Research - Working Paper Series*, 16045.

Rios-Avila, Fernando (2020): "Recentered influence functions (RIFs) in Stata: RIF regression and RIF decomposition." *The Stata Journal* 20(1): 51-94.

Rothe, Christoph (2015): "Decomposing the Composition Effect. The Role of Covariates in Determining Between-Group Differences in Economic Outcomes." *Journal of Business & Economic Statistics* 33(3): 323-337.


