---
output:
  github_document:
    toc: true
    toc_depth: 3
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# `ddecompose`: Detailed Distributional Decompositions

<!-- badges: start -->
[![R-CMD-check](https://github.com/samumei/ddecompose/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/samumei/ddecompose/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/samumei/ddecompose/branch/master/graph/badge.svg)](https://app.codecov.io/gh/samumei/ddecompose?branch=master)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![CRAN status](https://www.r-pkg.org/badges/version/ddecompose)](https://CRAN.R-project.org/package=ddecompose)
[![License: GPL (>= 3)](https://img.shields.io/badge/License-GPL%20%28%3E%3D%203%29-blue.svg)](https://choosealicense.com/licenses/gpl-3.0/)
<!-- badges: end -->


## Overview

The R package **ddecompose** implements the [Oaxaca (1973)](https://www.jstor.org/stable/2525981)-[Blinder (1973)](https://www.jstor.org/stable/144855) decomposition method and generalizations
of it that decompose differences in distributional statistics beyond the mean.

`ob_decompose()` decomposes differences in the mean outcome between two groups into one 
part explained by different covariates (composition effect) and into 
another part due to differences in the way covariates are linked to the 
outcome variable (structure effect). The function further divides the two
effects into the contribution of each covariate and allows for weighted 'doubly
robust' decompositions. For distributional statistics beyond the mean,
the function performs the RIF decomposition proposed by [Firpo, Fortin, and Lemieux (2018)]( https://doi.org/10.3390/econometrics6020028). 

`dfl_decompose()` divides differences in distributional statistics into an
composition effect and a structure effect using inverse probability
weighting as introduced by [DiNardo, Fortin, and Lemieux (1996)](https://www.jstor.org/stable/2171954). The function also allows
to sequentially decompose the composition effect into the contribution of
single covariates.

The package contains generic summary, print and plot functions for the results
and computes standard errors. This documentation provides a brief 
overview of the functions implemented in the package. For a more detailed 
discussion of decomposition methods including their respective assumptions and 
limitations, refer to [Fortin, Lemieux, and Firpo (2011)](https://economics.ubc.ca/wp-content/uploads/sites/38/2013/05/pdf_paper_nicole-fortin-decomposition-methods.pdf) and Firpo et al. (2018). 

## Installation 

You can hopefully soon install the CRAN version of `ddecompose`
```{r, eval=FALSE}
install.packages("ddecompose")
```

Until now, refer to the latest development version from GitHub:
```{r , eval=FALSE}
devtools::install_github("samumei/ddecompose")
```


## Background

### Oacaxa-Blinder Decomposition

The original decomposition method introduced by Oaxaca (1973) and Blinder (1973) 
divides the difference in the mean of an outcome variable (e.g., hourly wages)
between two groups $g = 0, 1$ into a part explained by differences in the 
mean of the covariates (e.g., educational level or experience) and into 
another part due to different linear regression coefficients (e.g., returns to education)
that link the covariates to the outcome variable. 

The method linearly models the relationship between the outcome $Y$ and covariates
$X$
$$Y_{g,i} = \beta_{g,0} + \sum^K_{k=1}X_{k,i}\beta_{g,k} + \varepsilon_{g,i},$$
where $\beta_{g,0}$ is the intercept and $\beta_{g,k}$ are the slope coefficients of
covariates $k = 1,\ldots, K$. Moreover, it is assumed that the error
term $\varepsilon$ is conditionally independent of $X$, i.e., 
$E( \varepsilon_{g,i} | X_{1,i}, \ldots ,X_{k,i}) = 0$, and that there is
an overlap in observable characteristics across groups ('common support'). 

The coefficients are estimated with OLS. Together with the sample means
of the covariates, one can derive a counterfactual mean  
$$\overline Y_C = \widehat \beta_{0,0} + \sum^K_{k=1}\overline X_{1,k} \widehat \beta_{0,k}$$
that would be observed if group $1$ had the same coefficients like group $0$. By 
adding and subtracting the counterfactual, the observed difference
$$\widehat\Delta^\mu_O = (\overline Y_1 - \overline Y_C) + (\overline Y_C - \overline Y_0) = \widehat\Delta^\mu_S + \widehat\Delta^\mu_C, $$

is divided into the aggregate structure effect 
$$\widehat\Delta^\mu_S  = (\widehat \beta_{1,0} - \widehat \beta_{0,0}) + \sum^K_{k=1}\overline X_{1,k}(\widehat \beta_{1,k} - \widehat \beta_{0,k}),$$


that captures outcome differences due to different coefficients, and the composition effect 
$$\widehat\Delta^\mu_X = \sum^K_{k=1} (\overline X_{1,k} - \overline X_{0,k})\widehat \beta_{0,k}, $$

which accounts for mean differences of the covariates. Note that
we could also combine the coefficients from group 1 with the covariates of group 
0 to define a counterfactual. Such a change in the "reference" group generally
leads to different results.

$\widehat\Delta^\mu_S$ and $\widehat\Delta^\mu_X$ denote the aggregate composition
terms. Since the method hinges on an additive linear model, the terms can be further
divided into the contribution of each covariate $k$ in a *detailed
decomposition*. For instance, the contribution of covariate $k = 1$ to the 
structure effect is $\overline X_{1,1}(\widehat \beta_{1,1} - \widehat \beta_{0,1})$, 
while $(\overline X_{1,1} - \overline X_{0,1})\widehat \beta_{0,1}$ is the covariate's
contribution to the composition effect.

### Reweighting Decomposition

The Oaxaca-Blinder decomposition is limited to the mean. Moreover, the 
decomposition terms are biased if the expectation of the outcome conditional
on the covariates is not linear (see [Barsky et al., 2002](https://www.jstor.org/stable/3085702)). DiNardo, Fortin, and Lemieux (1996), DFL hereafter, propose an alternative approach that overcomes both shortcomings. Instead of modeling the conditional mean, the method uses 
inverse probability weighting to define a counterfactual outcome distribution
that combines the conditional outcome distribution of one group and the covariates
distribution of the other group.  For instance, if we are interested in the 
outcomes of group 0 with covariates of group 1, we would reweight the outcome
distribution of group 0 such that its covariates distribution matches that of group 1
$$F_{Y_C}(y) = \int F_{Y_0}(y|x)dF_{X_1} (x)= \int F_{Y_0}(y|x)\Psi_X(x)dF_{X_0}(x).$$

By applying Bayes' rule, the reweighting factor,
$$\Psi_X(x) = \frac{dF_{X_1}(x)}{dF_{X_0}(x)} = \frac{P(g=0)P(g=1|x)}{P(g=1)P(g=0|x)},$$ 

can be expressed in terms of $P(g)$ and $P(g|x)$, the (conditional) probabilities of
belonging to group $g$. This allows us to estimate the reweighting factor using sample 
probabilities of each group and fitting conditional probability models (e.g.,
logit) in the joint sample. The estimated factors are then used to estimate weighted 
distributional statistics of interest (e.g., mean, quantiles or Gini coefficient) in 
the reference sample -- group 0 in the present example. The resulting counterfactual
distributional statistic, $\widehat\nu_C=\widehat\nu(F_{Y_C})$, is then contrasted 
with the observed difference
$$\widehat\Delta_O^{\nu} = (\widehat\nu_1 - \widehat\nu_C) + (\widehat\nu_C - \widehat\nu_0) = \widehat\Delta_S^\nu + \widehat\Delta_X^\nu,$$

which yields again an aggregate structure effect and aggregate composition effect.

The two decomposition terms account for the contribution of the covariates and of the
conditional outcome distribution, respectively, assuming common support and ignorability.
The latter condition asserts that the distribution of unobserved covariates $\varepsilon$ conditional on observed covariates $X$ is independent of group $g$.

### Sequential Decomposition

In contrast to the Oaxaca-Blinder decomposition, where the contributions of each
covariate simply add up, detailed decompositions are not straightforward in the
reweighting framework. However, DFL show that we can sequentially alter the covariates
distributions to decompose the composition effect into the contribution of single
covariates. For instance, assume we want to distinguish the effect of covariate $X_1$ 
(e.g., union status) from that of covariate $X_2$ (e.g., industry).
We begin again with the counterfactual distribution based on the conditional
outcome distribution of group 0 and the covariates distribution of group 1 
$$F_{Y_{C}}(y) = \iint F_{Y_0}(y|x_1,x_2)dF_{X_{1,1}}(x_1|x_2)dF_{X_{1,2}}(x_2)$$

and introduce a second counterfactual where we combine the 
conditional outcome distribution of group 0 as well as the conditional covariate 
distribution of $X_1$ given $X_2$ (e.g., union coverage by industry) of
group 0 with the covariates distribution $X_2$ (e.g., employment by industry) of group 1
$$F_{Y_{C,X_2}}(y) = \iint F_{Y_0}(y|x_1,x_2)dF_{X_{0,1}}(x_1|x_2)dF_{X_{1,2}}(x_2) $$

which can be expressed as the outcome distribution 0
$$F_{Y_{C,X_2}}(y) = \iint F_{Y_0}(y|x_1,x_2)dF_{X_{0,1}}(x_1|x_2)\Psi_{X_2}(x_2)dF_{X_{0,2}}(x_2),$$
reweighted by the factor
$$\Psi_{X_2}(x_2) = \frac{dF_{X_{1,2}}(x_2)}{dF_{X_{0,2}}(x_2)} =  \frac{P(g=0)P(g=1|x_2)}{P(g=1)P(g=0|x_2)}.$$

With the distributional statistics of the additional counterfactual, we can 
divide the aggregate decomposition effect into the contribution of each covariate
$$\widehat \Delta_X^{\nu} =  (\widehat \nu_C - \widehat \nu_{C,X_2}) + (\widehat \nu_{C,X_2} - \widehat \nu_0) = \widehat \Delta_{X_1}^\nu + \widehat \Delta_{X_2}^\nu.$$


Sequential decompositions are path dependent because the detailed composition
effects attributed to single covariates depend on the order of which we include the 
variables into the sequence. For instance, it matters if we reweight union coverage
by industry $F(x_1|x_2)$ or the industry employment given union coverage
$F(x_2|x_1)$.

Moreover, we get different results if we do not combine the marginal covariate 
distribution $X_2$ (e.g., industry employment) of group 1 with the conditional
distribution of $X_1$ given $X_2$ (e.g., union density by industry) of group 0 but
rather combine the marginal of group 0 with the conditional distribution of group 1
to derive the counterfactual, e.g.,
$$F_{Y_{C,X_1}}(y) = \iint F_{Y_0}(y|x_1,x_2)dF_{X_{1,1}}(x_1|x_2)dF_{X_0,2}(x_2).$$
where we would reweight group 0 with a slightly different factor 
$$\Psi_{X_1}(x_1,x_2) = \frac{dF_{X_{1,1}}(x_1|x_2)}{dF_{X_{0,1}}(x_1|x_2)} =  \frac{P(g=0|x_2)P(g=1|x_1,x_2)}{P(g=1|x_2)P(g=0|x_1,x_2)}.$$

### 'Doubly Robust' Oaxaca-Blinder Decomposition

A robust and path independent alternative for detailed decompositions at the mean 
is to combine DFL reweighting with the linear Oaxaca-Blinder method (see 
Fortin et al., 2011: 48-51). This approach has the valuable side effect of 
accounting for potential errors introduced by an incomplete inverse probability 
weighting and the linear model specification, respectively.

The estimation proceeds in two steps. First, the reweighting function $\widehat\Psi_X(x)$, which matches the characteristics of group $0$ (if group 0 is the reference group) to those of group $1$, is derived. Second, the linear model and the covariate means are estimated in the two observed samples as well as in the reweighted sample $0$, yielding $\widehat \beta_{C,0}$, $\widehat \beta_{C,k}$, and $\overline X_{C,k}$.

The mean of the reweighted sample builds the main counterfactual to derive the aggregate structure and composition effects, respectively. The detailed decomposition terms are also evaluated with respect to the statistics of the reweighted sample, i.e., 

$$\widehat\Delta^\mu_{O,R} = (\widehat \beta_{1,0} - \widehat \beta_{C,0}) + \sum^K_{k=1} (\overline X_{1,k}\widehat \beta_{1,k} - \overline X_{C,k}\widehat \beta_{C,k}) + \sum^K_{k=1} (\overline X_{C,k}\beta_{C,k} - \overline X_{0,k}\widehat \beta_{0,k}) = \widehat\Delta^\mu_{S,R}  + \widehat\Delta^\mu_{X,R}.$$
These decomposition terms can be further decomposed into a structure and into a composition effect, respectively. This separates the errors from reweighting $\widehat\Delta^\mu_{S,e}$ and the linear specification $\widehat\Delta^\mu_{X,e}$, respectively, and yields a "pure" composition effect $\widehat\Delta^\mu_{X,p}$ and a "pure" structure effect $\widehat\Delta^\mu_{S,p}$ for each covariate. 

Specifically, the structure effect can be written as

$$\widehat\Delta^\mu_{S,R} = (\widehat \beta_{1,0} - \widehat \beta_{C,0}) + \sum^K_{k=1}\overline X_{1,k}(\widehat \beta_{1,k} - \widehat \beta_{C,k}) + \sum^K_{k=1} (\overline X_{1,k} - \overline X_{C,k})\widehat \beta_{C,k} = \widehat\Delta^\mu_{S,p} + \widehat\Delta^\mu_{S,e}.$$

By comparing $\beta_{1,k}$ to the coefficients of the reweighted group $0$, $\widehat \beta_{C,k}$, the true underlying structure effect can be identified. Moreover, the reweighting error, $\widehat\Delta^\mu_{S,e}$, indicates how accurately the reweighting function $\widehat\Psi(X_i)$ reweights the covariates distribution of group $0$ to that of group $1$. Thus, identifying the reweighting error allows to measure the "pure" structure effect, $\widehat\Delta^\mu_{S,p}$. If the composition of group $1$ is equal to the reweighted group $0$, the reweighting error is zero.

Similarly, the additional decomposition of the composition effect reads as

$$\widehat\Delta^\mu_{X,R} = \sum^K_{k=1} (\overline X_{C,k} - \overline X_{0,k})\widehat \beta_{0,k} + (\widehat \beta_{C,0} - \widehat \beta_{0,0}) + \sum^K_{k=1}\overline X_{C,k}(\widehat \beta_{C,k} - \widehat \beta_{0,k}) = \widehat\Delta^\mu_{X,p} + \widehat\Delta^\mu_{X,e}.$$

The specification error, $\widehat\Delta^\mu_{X,e}$, measures the extent to which the estimated coefficients change due to a different distribution of covariates if the (wage) structure remains the same. Thereby, the "pure" effect, $\overline X_{C,k} - \overline X_{0,k}$, for each covariate can be estimated, summing to the aggregate pure structure effect, $\widehat\Delta^\mu_{X,p}$. If the model is truly linear, i.e., correctly specified, $\beta_C$ of the reweighted group $0$ will be identical to $\beta_0$ and the specification error will be zero (Fortin et al. 2011, p. 49-50). 

The reweighted OB decomposition is "doubly robust" as it yields consistent estimates
even if either the linear model or the reweighting estimator is misspecified. In
contrast to the simple reweighting or linear approach it does not hinge on a single 
correctly specified model. While the reweighted OB decomposition is doubly robust 
and path independent, it is limited to the mean.


### Reweighted RIF Regression Decomposition 

A path independent method that goes beyond the mean is the RIF decomposition 
of Firpo, Fortin, and Lemieux (2018). The approach approximates the expected value
of the 'recentered influence function' (RIF) of the distributional statistic (e.g., quantile, variance, or Gini coefficient) of an outcome variable conditional on covariates with linear regressions. RIF regression coefficients can be consistent estimates of the marginal effect 
of a small change in the expected value of a covariate to the distributional statistics of 
an outcome variable (see [Firpo et al., 2009a](https://www.econometricsociety.org/publications/econometrica/2009/05/01/unconditional-quantile-regressions) and the documentation of the companion package [`rifreg`](https://github.com/samumei/rifreg)). Thus, they can be used to decompose between-group difference in distributional statistics. Firpo et al. (2018) combine the RIF regressions again 
with the reweighting estimator to avoid specification errors. 

First, the approach computes the recentered influence function (RIF) of the outcome variable $Y$ and a distributional statistic of interest $\nu$. Then, an OLS regression of the transformed outcome variable $Y$ is run on the explanatory variables $X$. Thereafter, the decomposition method is analogous to the original OB method:

$$\widehat\Delta^\nu_{O,R} =  \widehat\Delta^\nu_{S,p} + \widehat\Delta^\nu_{S,e} + \widehat\Delta^\nu_{X,p} +  \widehat\Delta^\nu_{X,e}.$$ 
With $\widehat\Delta^\nu_{S,p}$ and $\widehat\Delta^\nu_{S,e}$ estimating the pure structure effect and the reweighting error as

$$\widehat\Delta^\nu_{S,R} =  \widehat\Delta^\nu_{S,p} + \widehat\Delta^\nu_{S,e} = (\widehat \beta_{1,0} - \widehat \beta_{C,0}) + \sum^K_{k=1}\overline X_{1,k}(\widehat \beta_{1,k} - \widehat \beta_{C,k}) + \sum^K_{k=1} (\overline X_{1,k} - \overline X_{C,k})\widehat \beta_{C,k}$$
and $\widehat\Delta^\nu_{X,p}$ and $\widehat\Delta^\nu_{X,e}$ estimating the pure coefficient effect and the specification error: 

$$\widehat\Delta^\nu_{X,R} = \widehat\Delta^\nu_{X,p} + \widehat\Delta^\nu_{X,e} = \sum^K_{k=1} (\overline X_{C,k} - \overline X_{0,k})\widehat \beta_{0,k} + (\widehat \beta_{C,0} - \widehat \beta_{0,0}) + \sum^K_{k=1}\overline X_{C,k}(\widehat \beta_{C,k} - \widehat \beta_{0,k})$$
with the RIF regression coefficients $\widehat \beta$ and the covariate means $\overline X$ of groups
$0$, $1$, and $C$, the reweighted reference group 0, respectively. 

Again, the specification error increases if the conditional expectation of the RIF
is not well approximated by the linear model. Moreover, as the RIF regression coefficients
capture the marginal effects of small location shifts in the covariates distribution on the
distributional statistic of the outcome variable, the specification error can
be large if the location shifts are substantial or if the between-group differences 
in the covariates distribution relate to higher moments or the dependence structure 
(see also [Rothe, 2012: 16-19](https://docs.iza.org/dp6397.pdf) or [Rothe, 2015: 328](https://doi.org/10.1080/07350015.2014.948959)).


### Inference

Analytical standard errors are straightforward for the original OB decomposition
under the assumption of independence between the groups (see Jann, [2005](https://boris.unibe.ch/69506/1/oaxaca_se_handout.pdf) and [2008](https://journals.sagepub.com/doi/abs/10.1177/1536867X0800800401)). [Firpo (2007)](https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1468-0262.2007.00738.x) and 
[Firpo and Pinto (2016)](https://www.jstor.org/stable/26609621) develop asymptotic theory
for the reweighting estimator. [Firpo, Fortin, and Lemieux  (2009b)](https://www.econometricsociety.org/publications/econometrica/2009/05/01/unconditional-quantile-regressions/supp/6822_extensions_0.pdf) and Firpo et al. (2018) do the same for the RIF estimator and for the RIF decomposition terms, respectively. The authors propose to bootstrap standard errors. `ddecompose` allows to bootstrap standard errors in both `ob_decompose()` and `dfl_decompose()`. For the classical OB decomposition, analytical standard errors are available, too.


## Examples 

The following examples illustrate the workings of the main decomposition functions in `ddecompose`. We use a sample of the National Longitudinal Survey (NLSY) 79 containig wage data from the year 2000 of workers who were aged 35 to 43 in that year. The data is from O'Neill and O'Neill (2006) and is used as an illustration of the Oxaca-Blinder mean decomposition in Fortin, Lemieux, and Firpo (2011). The data contains 2655 male and 2654 female observations, respectively.

```{r, eval = TRUE} 
library(ddecompose)
data("nlys00")
```


### Oaxaca-Blinder Decomposition 

Let us decompose the gender wage gap into composition and structure effect, using the standard Oaxaca-Blinder decompostion without reweighting. First, we specify a wage structure model and then run the estimation. 

```{r, eval = TRUE}

model <- log(wage) ~  age + region + education + 
                        years_worked_civilian + years_worked_military + 
                        part_time + family_responsibility + industry 

gender_gap_decomposition <- ob_decompose(formula = model,
                                        data = nlys00,
                                        group = female)
```

In the default settings, heteroscedasticity-consistent standard errors are calculated and factor variables are not normalized. Group 0 is defined as the reference group (the group with the lower factor level) and is subtracted from group 1. Using `summary()`, we can display the decompostion formula and the estimation results.


```{r, eval=TRUE}
summary(gender_gap_decomposition)
```





`ddecompose` comes with a convenient plot function. To plot the overall composition and structure effect, we have to set `detailed_effects = FALSE`.

```{r, eval=TRUE}
plot(gender_gap_decomposition, detailed_effects = FALSE)
```


To change the reference group, we `reference_0` needs to be set to `FALSE`. Note that this does not change the subtraction terms in the decomposition, i.e.,  (X1 - X0) and (b1 - b0). To change the direction of the subtraction, the parameter `subtract_1_from_0` needs to be set to `TRUE`. This merely changes the sign of the decomposition results. We can indicate the weights column of the `data` in the `weights` parameter and normalize factor variables by setting `normalize_factors = TRUE`. 


### 'Doubly Robust' Decompostion 
To compute a more robust decomposition, we can add reweighting to the decompostion. Thereby, we also estimate the specification and reweighting error of the results. By default, the same function model is used for reweighting as for the decompostion. However, it is advisable to add a more flexible reweighting model, taking interactions into account. The reweighting formula is added to the decomposition formula, separated by a `|`. For this decompostion method, standard errors need to be computed using bootstrap. By increasing the number of `cores` the bootstrap replications are computed on, we can reduce computation time. Per default, 100 bootstrap replications are calculated. 



```{r, eval = TRUE}

model_w_reweighting <- log(wage) ~  
                           age + region + education + 
                           years_worked_civilian + years_worked_military + 
                           part_time + family_responsibility + industry |
                           age + region + education + 
                           years_worked_civilian + years_worked_military + 
                           part_time + family_responsibility + industry +
                           education:region + age:education 



gender_gap_decomposition_w_reweighting <- 
    ob_decompose(formula = model_w_reweighting,
                 data = nlys00,
                 group = female, 
                 reweighting = TRUE, 
                 bootstrap = TRUE, 
                 cores = 4)
```


The default method to fit and predict conditional probabilities, used to derive the reweighting factor, is a logit model. However, one can also use `reweighting_method = "fastglm"`, for a faster logit model computation or `random forests` for a different reweighting method. If `trimming` is set to `TRUE`, observations with dominant reweighting factor values are trimmed. Per default, reweighting factor values are trimmed according to rule of Huber, Lechner, and Wunsch (2013). Therby, the `trimming_threshold`, i.e., the maximal accepted relative weight of the reweighting factor value (inverse probability weight) of a single observation, is set to `sqrt(N)/N`, where `N` is the number of observations in the reference group. The trimming threshold can also be set manually to a numeric value. 

We can aggregate the detailed effects in the `summary()` and `plot()` function. For instance, if we want to separate personal and contextual factors that explain the wage gap, we can aggregate these variables in a list.

```{r, eval=TRUE}

custom_aggregation <- list(`Personal Factors` = c("age", 
                                                  "education<10 yrs", 
                                               "educationHS grad (diploma)",
                                               "educationHS grad (GED)",
                                               "educationSome college",
                                               "educationBA or equiv. degree",
                                               "educationMA or equiv. degree",
                                               "educationPh.D or prof. degree",
                                               "part_time",
                                               "family_responsibility"), 
                           `Experience` = c("years_worked_civilian",
                                            "years_worked_military"), 
                           `Contextual Factors` = c("regionNorth-central",
                                                 "regionSouth",
                                                 "regionWest",
                                                 "industryManufacturing",
                                                 "industryEducation, Health, Public Admin.",
                                                 "industryOther services"))

plot(gender_gap_decomposition_w_reweighting, 
     custom_aggregation = custom_aggregation)
```

If we add reweighting, the `plot()` and `summary()` function also display the Specification and Reweighting Error. To reduce the reweighting error, a more flexible reweighting model is advisable. 

### RIF Regression Decomposition

To decompose group differences beyond the mean with `ob_deco` we are using RIF regressions. In the following examples, we will analyze the changes in wage inequality between 1983/85 and 2003/05 and assess what covariates contribute to explaining these changes. First, we look at changes in the variance. Then, we decompose the wage gap at each decile. The data used is a sample of the the Merged Outgoing Rotation Group of the Current Population Survey from 1983 to 1985 and 2003 to 2005, respectively, used as example by Fortin, Lemieux & Firpo  (2011). 


```{r, eval = TRUE}

data("men8305")

model_rifreg <- log(wage) ~ union*(education + experience) + education*experience 

# Variance
variance_decomposition <- ob_decompose(formula = model_rifreg,
                 data = men8305,
                 group = year, 
                 reweighting = TRUE, 
                 rifreg_statistic = "variance",
                 bootstrap = TRUE, 
                 cores = 4)

print(variance_decomposition)


# Deciles
deciles_decomposition <- ob_decompose(formula = model_rifreg,
                 data = men8305,
                 group = year, 
                 reweighting = TRUE, 
                 rifreg_statistic = "quantiles",
                 rifreg_probs = c(1:9)/10,
                 bootstrap = TRUE, 
                 cores = 4)

plot(deciles_decomposition)


```

The RIF functions for the following functions are currently implemented: `"quantiles"`, `"mean"`, `"variance"`, `"gini"`, `"interquantile_range"`, and `"interquantile_ratio"`. However, `ob_decompose` also allows to pass a custom RIF function for the decomposition, by setting `rifreg_statistic = "custom"` and passing the custom function to `custom_rif_function`. [Cowell and Flachaire (2007)](https://doi.org/10.1016/j.jeconom.2007.01.001), [Essama-Nssah & Lambert (2012)](https://doi.org/10.1108/S1049-2585(2012)0000020009), and [Rios-Avila (2020)](https://doi.org/10.1177/1536867X20909690) derive the influence functions for an array of distributional statistics. More information about RIF regressions can be found in the documentation of the companion package [`rifreg`](https://github.com/samumei/rifreg).  

Custom RIF functions must specify a `dep_var` parameter for the outcome variable $Y$, a `weights` for potential sample weights, and `probs`. If they are not needed, they must be set to `NULL` in the function definition (e.g. `probs = NULL`).

The following example shows how to write the RIF for the top 10 percent income share and, then, to estimate the RIF regression decomposition using this custom function. The formula for this specific RIF can be found in Essam-Nssah & Lambert (2012) or Rios-Avila (2020).

```{r, eval = TRUE}

# custom RIF function for top 10% percent income share
custom_top_inc_share <- function(dep_var,
                                 weights,
                                 probs = NULL,
                                 top_share = 0.1) {
  top_share <- 1 - top_share
  weighted_mean <- weighted.mean(
    x = dep_var,
    w = weights
  )
  weighted_quantile <- Hmisc::wtd.quantile(
    x = dep_var,
    weights = weights,
    probs = top_share
  )
  lorenz_ordinate <- sum(dep_var[which(dep_var <= weighted_quantile)] *
    weights[which(dep_var <= weighted_quantile)]) /
    sum(dep_var * weights)
  if_lorenz_ordinate <- -(dep_var / weighted_mean) * lorenz_ordinate +
    ifelse(dep_var < weighted_quantile,
      dep_var - (1 - top_share) * weighted_quantile,
      top_share * weighted_quantile
    ) / weighted_mean
  rif_top_income_share <- (1 - lorenz_ordinate) - if_lorenz_ordinate
  rif <- data.frame(rif_top_income_share, weights)
  names(rif) <- c("rif_top_income_share", "weights")
  return(rif)
}

custom_decomposition <- ob_decompose(formula = model_rifreg,
                 data = men8305,
                 group = year, 
                 reweighting = TRUE, 
                 rifreg_statistic = "custom",
                 custom_rif_function = custom_top_inc_share,
                 bootstrap = TRUE, 
                 cores = 4)

plot(custom_decomposition)

```




### Reweighting Decomposition

Now, we present the usage of the other main function: `dfl_decompose()`. Fortin, Lemieux, and Firpo (FLF, 2011, p. 79-88) decompose the increase in USmale wage inequality between the early 1980s and the early 2000s using the CPS data. In this example, we replicate their results and treat the observations from
1983 to 1985 as reference group that is reweighted.

```{r, eval = FALSE}
library(ddecompose)
data("men8305")

flf_model <- log(wage) ~ union*(education + experience) + education*experience
flf_male_inequality  <- dfl_decompose(flf_model, 
                                  data = men8305,
                                  weights = weights,
                                  group = year, 
                                  bootstrap = TRUE, 
                                  cores = 4)
```
We can summarize the results:
```{r, eval = FALSE}
summary(flf_male_inequality)
```

Using `plot()`, we can illustrate the decomposition accross different quantiles.
```{r, eval = FALSE}
plot(flf_male_inequality)
```




## Replication of Firpo, Fortin, and Lemieux (2018)

To validate the functions and provide users with an additional example, we replicate the reweighted RIF regression decomposition estimates in Firpo et al. (2018, p. 30). In their empirical example, Firpo et al. focus in changes in wage inequality in the US between 1988 and 2016. Using a large sample of male log wages in 1988-1990 (268'494 observations) and in 2014-2016 (236'296 observation) based on the Outgoing Rotation Group (ORG) supplement of the Current Population Survey (CPS), they attribute changes in variance, gini, and interquantile ranges between the two time periods to changes in composition (e.g., union coverage, education, and experience) and changes in wage structure. 

This replication follows the Stata replication code, that the authors published alongside the paper and that is available on their website. (INSERT LINK). The highest wages in the public CPS data set are unavailable due to privacy concerns. The Firpo et al. impute these wages using a random uniform Pareto distribution (CHECK WORDING). Since random numbers are generated differently in R and Stata, even with an equivalent seed. Therefore, we perform the entire data preparation in Stata up until the estimation of the decomposition, i.e., the "oaxaca" commands in Stata. This ensures that changes in the estimation results between our function and the orignal paper do not stem from different input data. 

To reproduce the code below, make sure you download the entire data set and Stata code from the journal’s website, execute the code up to the oaxaca command, save the data and load it into your R environment. 


### Loading Data


```{r, eval=FALSE}
# Make sure you execute the Stata Code until the oaxaca commands 
# and save the data in the appropriate folder.
men8816_t4 <- readstata13::read.dta13("data-raw/usmen8816_t4.dta")

# Removing redundant observations - we replicate the reweighting within the function
men8816_t4 <- men8816_t4[men8816_t4$time <= 1,] 
```


### Reweighted RIF Regression Decomposition (Table 4)

The model is specified as in the Stata files, containing weights, computing bootstrapped standard errors with 100 iterations, and setting a fixed bandwidth of 0.06 for the epanechnikov kernel density estimation.

```{r eval=FALSE}

set.seed(987421)

ffl_model_with_reweighting <- as.formula(
    paste("lwage2 ~ covered + nonwhite + nmarr + 
          ed0 + ed1 + ed3 + ed4 + ed5 + ",
          paste(grep(paste0("^ex(", paste(c(1:4, 6:9), collapse = "|"), ")$"), 
                     names(men8816_t4), value = T), collapse = " + "), " + ",
          paste(grep(paste0("^occd(", paste(c(11:60, 80:91), collapse = "|"), ")$"),
                     names(men8816_t4), value = T), collapse = " + "), " + ",                     
          paste(grep(paste0("^indd(", paste(c(1, 3:14), collapse = "|"), ")$"), 
                     names(men8816_t4), value = T), collapse = " + "), " + pub | ",
          paste("covered + nonwhite +",
                paste(grep("^marr", names(men8816_t4), value = TRUE), collapse = " + "), "+",
                paste(c("ed0", "ed1", "ed3", "ed4", "ed5"), collapse = " + "), "+",
                paste(grep("^ex[1-4]|^ex[6-9]", names(men8816_t4), value = TRUE), collapse = " + "), "+",
                paste(grep("^uned", names(men8816_t4), value = TRUE), collapse = " + "), "+",
                paste(grep("^unex", names(men8816_t4), value = TRUE), collapse = " + "), "+",
                paste(grep("^ex[1-9]ed", names(men8816_t4), value = TRUE), collapse = " + "), "+",
                paste(grep("^pub", names(men8816_t4), value = TRUE), collapse = " + "), "+",
                paste(grep("^indd(1|1e|[3-9]|10|11|13|14)(?!2)", 
                           names(men8816_t4), perl = TRUE, value = TRUE), collapse = " + "), "+",
                paste(grep("^occd", names(men8816_t4), value = TRUE), collapse = " + "))))


# Interquantile Ratio 90-10 
decompose_90_10  <- ob_decompose(formula = ffl_model_with_reweighting,
                                data = men8816_t4,
                                weights = eweight,
                                group = time,
                                reference_0 = TRUE,
                                rifreg_statistic = "interquantile_range",
                                rifreg_probs = c(0.9, 0.1),
                                bw = 0.065,
                                kernel = "epanechnikov",
                                reweighting = TRUE,
                              reweighting_method = "fastglm",
                                trimming = TRUE,
                                trimming_threshold = 100,
                                bootstrap = TRUE,
                                bootstrap_iterations = 100)

## Variance
decompose_variance  <- ddecompose::ob_decompose(formula = ffl_model_with_reweighting,
                                 data = men8816_t4,
                                 weights = eweight,
                                 group = time,
                                 reference_0 = TRUE,
                                 rifreg_statistic = "variance",
                                 reweighting = TRUE,
                              reweighting_method = "fastglm",
                                 trimming = TRUE,
                                 trimming_threshold = 100,
                                 bootstrap = TRUE,
                                 bootstrap_iterations = 100)

## Gini 

# Updating the model
gini_model_raw <- update(ffl_model_with_reweighting, exp(.) ~ .)
gini_model_character <- as.character(gini_model_raw)
gini_model_split <- strsplit(gini_model_character, "~")
ffl_model_with_reweighting_gini <- as.formula(paste(gini_model_split[[2]], "~",
                                                     gsub("\\(|\\)", "", gini_model_split[[3]][1])))

set.seed(130234976)
decompose_gini  <- ddecompose::ob_decompose(formula = ffl_model_with_reweighting_gini,
                               data = men8816_t4,
                               weights = eweight,
                               group = time,
                               rifreg_statistic = "gini",
                               reweighting = TRUE,
                              reweighting_method = "fastglm",
                               trimming = TRUE,
                               trimming_threshold = 100,
                               bootstrap = TRUE,
                               bootstrap_iterations = 100, 
                              cores = 1)




```



### Presentation of Results

Describe the results and to what extent they are similar to those of ffl and where they differ and why. 

Also note that the replication with the same imputed data and code to replicate table 1-4 is available upon request. 

MAKE SURE THAT YOU write down where the differences are arising: 

- Data (reweighting)
- seed
- rounding
- density estimation 
- SE (includes Reweighting)

MAYBE ALSO NOTE HOW MUCH THE DIFFERENCE IS, IN THE OTHER TABLES ETC. AND THAT THESE FILES ARE AVAILABLE UPON REQUEST BZW: IN THE TEST_FILE XXX -> PUT IN A SEPARATE TEST FILE!



```{r eval=TRUE}



```






Looking at the plots, we see that our plots correspond to those presented by Firpo, Fortin, and Lemieux (2009a: 965). The plot example illustrates, how the plot from the generic rifreg::plot() function can be further enhanced, for instance with a horizontal line indicating the OLS coefficient for comparison.

This validation example illustrates that the rifreg package works as intended in computing RIF regressions and reliably yields the expected results.


## Credits

David Gallusser & Samuel Meier

## References

Barsky, Robert, John Bound, Kerwin Kofi Charles, and Joseph P. Lupton. 2002. “Accounting for the Black-White Wealth Gap: A Nonparametric Approach.” *Journal of the American Statistical Association* 97: 663–73.

Cowell, Frank A., and Emmanuel Flachaire. 2007. "Income distribution and inequality measurement: The problem of extreme values." *Journal of Econometrics* 141: 1044–1072.

Essama-Nssah, Boniface, and Peter J. Lambert. 2012. “Influence Functions for Policy Impact Analysis.” In John A. Bishop and Rafael Salas, eds., *Inequality, Mobility and Segregation: Essays in Honor of Jacques Silber*. Bigley, UK: Emerald Group Publishing
Limited.

Firpo, Sergio. 2007. "Efficient Semiparametric Estimation of Quantile Treatment Effects." *Econometrica* 75(1): 259-276.

Firpo, Sergio, Nicole M. Fortin, and Thomas Lemieux. 2007a. “Unconditional Quantile Regressions.” *Technical Working Paper 339, National Bureau of Economic Research*. Cambridge, MA. 

Firpo, Sergio, Nicole M. Fortin, and Thomas Lemieux. 2009a. "Unconditional Quantile Regressions." *Econometrica* 77(3): 953-973.

Firpo, Sergio, Nicole M. Fortin, and Thomas Lemieux. 2009b. "Supplement to 'Unconditional Quantile Regressions'." *Econometrica Supplemental Material*, 77.

Firpo, Sergio, Nicole M. Fortin, and Thomas Lemieux. 2018. "Decomposing Wage Distributions Using Recentered Influence Function Regressions." *Econometrics* 6(2):28.

Fortin, Nicole M., Thomas Lemieux, and Sergio Firpo. 2011. "Decomposition Methods in Economics." *National Bureau of Economic Research - Working Paper Series*, 16045.

Firpo, Sergio, and Pinto, Christine. 2016. "Identification and Estimation of Distributional Impacts of Interventions Using Changes in Inequality Measures." *Journal of Applied Econometrics*, 31(3), 457–486. 

Huber, Martin, Michael Lechner, and Conny Wunsch. 2013. "The performance of estimators based on the propensity score." *Journal of Econometrics* 175(1): 1-21.

Jann, Ben. 2005. "Standard errors for the Blinder-Oaxaca decomposition." *3rd German Stata Users’ Group Meeting 2005*. Available from [https://boris.unibe.ch/69506/1/oaxaca_se_handout.pdf](https://boris.unibe.ch/69506/1/oaxaca_se_handout.pdf).

Jann, Ben. 2008. "The Blinder–Oaxaca Decomposition for Linear Regression Models". *Stata Journal* 8: 435–479.

Kline, Pat. 2009. "Blinder-Oaxaca as a Reweighting Estimator". *UC Berkeley mimeo.*

Rios-Avila, Fernando. 2020. "Recentered influence functions (RIFs) in Stata: RIF regression and RIF decomposition." *The Stata Journal* 20(1): 51-94.

Rothe, Christoph. 2012. "Decomposing the Composition Effect. The Role of Covariates in Determining Between-Group Differences in Economic Outcomes." *IZA Discussion Paper* No. 6397.

Rothe, Christoph. 2015. "Decomposing the Composition Effect. The Role of Covariates in Determining Between-Group Differences in Economic Outcomes." *Journal of Business & Economic Statistics* 33(3): 323-337.



